Bayesian Statistical Modeling with Predictors from LLMs

### 2024-03-29 04-paper/paper-LLM-probmod.tex
updated §5, added LLaMA results intepretation

To address this question, we assessed item-level scores $\text{S}_{ki}$ from the text-davinci-003 instance GPT-3.5 August 2023 \citep{BrownMann2020:Language-Models} for a text-based version of the reference game from Section~\ref{experiment-reference-games}.
% \footnote{
  % Numerically, we worked with \emph{average} log-probabilities. As in our experiment, all choice options had the same number of words, this is largely irrelevant, except for the effect of the $\alpha$ parameter in the interpretation condition. Using length-corrected log-probabilities is the same as using raw log-probabilities with an $\alpha$ parameter divided by the number of words. In this way, we make the $\alpha$ parameter more meaningfully comparable across conditions.}
The task description $x_{k}$ for item $I_{k}$ is a text supplied as prompt (see Appendix~\ref{sec:examples-items-llm} for an example).
The choice options are categorized as in the human experiment.\footnote{
  In the current set-up the response type ``distractor'' has two instantiations in the production condition. Since choice options are a single word in the production condition, for simplicity, we lump both of the distractor words together and treat them as a single option by taking the sum of the log-probabilities for both distractor words.}
Similar to the RSA model fit, we allow for random errors with probability $\epsilon_{C}$:
%
\begin{align*}
  P_{\text{item}}^{\text{LLM}}\left ( y_{ki} \mid C; \alpha_{c}, \epsilon_{c} \right )
  = \epsilon_{c} \  P_{\text{item}}^{\text{LLM}}\left ( y_{ki} \mid C; \alpha_{c}, \epsilon_{c} \right ) \ \ + \ \ \nicefrac{\epsilon_{c}}{3}   \,.
\end{align*}
%
The resulting likelihood function for item-level data with item-level predictions is, where priors on parameters are as for the RSA model described in Section~\ref{sec:model-pred-from}.:
%
\begin{align*}
 P^{\text{LLM}}_{\text{item}}(D_{c} \mid C; \alpha_{c}, \epsilon_{c}) = \prod_{k = 1}^{m} \text{Multinomial} \left ( D_{c}^{k}, \tuple{P^{\text{LLM}}_{\text{item}}(y_{ki} \mid C;  \alpha_{c}, \epsilon_{c})}_{1 \le l \le 3} \right )\,.
\end{align*}
%

@@

While LLMs do not make conditional-level predictions as such, they can be derived from item-level scores $\text{S}_{ki}$ by averaging over all items belonging to the relevant condition.
There are many ways of averaging item-level information.
Figure~\ref{fig:measures-overview} shows three salient approaches, which differ in what the underlying item-level measure is: the raw scores $\text{S}_{ki}$, the item-level probability derived from it (as introduced in Section~\ref{motivation} and used in Section~\ref{sec:item-level-pred}), or the predictions from the winner-takes-all strategy commonly used in benchmark testing (see Section~\ref{motivation}).

@@

The \emph{average-scores predictor} first aggregates the item-level scores, and then transposes the averages into (scaled) probabilities using the usual parameterized softmax function:\footnote{
  The reported results average over the multi-set \(\set{I_1, \dots I_m}\) of items that occurred in the human experiment for condition $C$.
  By using a multi-set, which may contain a single item multiple times, we produce aggregate predictions for exactly the set of items that the participant group saw, which provides the most fitting counterpart to the human data.}
%
\begin{align*}
  & P_{\text{cond}}^{\text{SCR}}(r_{i} \mid C, \alpha_{c})
    \propto \expo \left [  \alpha_{c} \ \frac{1}{m} \ \sum_{k=1}^{m} \text{S}_{ki}  \right ] \,.
    \tag*{\textcolor{gray}{[average scores (narrow-scope aggregation)]}}
\end{align*}
%
The \emph{average-probabilities predictor} first transposes scores into probabilities with a parameterized softmax function, and only aggregates over items last:
\begin{align*}
  % & \textbf{wide-scope aggregation} \\
  & P_{\text{cond}}^{\text{PRB}}(r_{i} \mid C, \alpha_{c})
    = \frac{1}{m} \ \sum_{k=1}^{m} P_{\text{item}}^{\text{LLM}}(y_{ki} \mid C, \alpha_{c}) \,.
    \tag*{\textcolor{gray}{[average probabilities (wide-scope aggregation)]}}
\end{align*}
%
Finally, the \emph{average-WTA predictor} considers the prediction of the WTA strategy as the basic item-level unit to aggregate over.
To add parameterized scaling to this method, a power-law transformation is a natural choice:
\begin{align*}
  % & \textbf{intermediate-scope aggregation} \\
  & P_{\text{cond}}^{\text{WTA}}(r_{i} \mid C, \alpha_{c})
    \propto  \left [ \frac{1}{m} \ \sum_{k=1}^{m} P_{\text{item}}^{\text{WTA}}(y_{ki} \mid C) \right ]^{\alpha_{c}}\,,
    \tag*{\textcolor{gray}{[average WTA (intermediate-scope aggregation)]}}
\end{align*}
where $P_{\text{item}}^{\text{WTA}}(y_{ki} \mid C) = \lim_{\alpha} P_{\text{item}}^{\text{LLM}}(y_{ki} \mid C, \alpha)$.
\bigskip

The \emph{average-scores} and \emph{average-probabilities} predictions are equivalent if there is only one item, in which case the prediction of the \emph{average-WTA} method is the special case of $\alpha \rightarrow \infty$.
For cases with more than one item, the predictions of the three predictors are not guaranteed to be the same.
Conceptually, the \emph{average-probabilities} and the \emph{average-WTA} predictors, but not the \emph{average-scores} predictor, are compatible with a picture in which condition-level predictions result from the actual predictions at the item level.
The difference between the \emph{average-probabilities} and the \emph{average-WTA} predictor is that the latter is an aggregate of an item-level predictor that is demonstrably bad, as shown in the previous Section~\ref{sec:item-level-pred}.

Using the same approach as for the RSA model in Section~\ref{sec:model-pred-from}, we can build likelihood functions for condition-level data around the three predictors introduced above.
With the same priors and methods used before, we obtain samples from the posterior over parameters and samples from the posterior predictive distributions.
Summary statistics for posteriors over model parameters are shown in Table~\ref{tab:sumStats}.
What is noteworthy is that for the \emph{average-WTA} model, the estimates of $\alpha_{c}$ in the production condition do not rule out, in fact lie close to, the special value $\alpha_{c}=1$, for which the power-law transformation is the identity function.
This means that just averaging WTA-responses at the item-level yields a reasonable predictor for the production data at the condition-level.
However, for the interpretation condition, the value $\alpha_{c}=1$ is clearly outside the range of credible parameter values, so that a simple recipe like ``always average WTA-responses without transformation'' is not a viable strategy for good condition-level predictions in general.
It is also worth noting that for the \emph{average-probability} model values of $\alpha_{c}$ close to one are virtually indistinguishable from the item-level predictions of the WTA strategy (see Figure~\ref{fig:closeness-target-item-level}).
This suggests that, for the production data, aggregation of item-level predictions from the WTA-strategy gives good condition-level predictions.
Since the production data is also very biased towards almost exclusively target choices by human participants, this is not too surprising.

Figure~\ref{fig:refgame-counts} shows the summary statistics (means and 95\% credible intervals) for each model's posterior predictive distribution.
We find that only the theoretical model (RSA) and the \emph{average-WTA} model pass this ``visual posterior predictive check'' for both conditions; the other two models both overpredict the target choice rate and underpredict the competitor choice rate in the interpretation condition.
To corroborate the visual impression, Table~\ref{tab:sumStats} shows sample-based estimates of Bayesian posterior predictive $p$-values, using likelihood of the observed data as a test statistics.
Consequently, the results from Table~\ref{tab:sumStats} suggest that the \emph{average-scores} model fails on the interpretation data, and is at most borderline compatible with the production data; that the \emph{average-probabilities} model is able to reproduce the production data, but fails on the interpretation data; and that only the \emph{average-WTA} model does not fail to capture the data from both conditions.

These results tell us that not all ways of deriving condition-level predictions by averaging over item-level variation are equally good.
Some approaches clearly fail basic checks for statistical goodness-of-fit.
On the positive side, we also find that there is at least one model with predictors based on LLM-measures, namely the \emph{average-WTA} model, which is able to recover the patterns in the data.
In other words, there is a way of deriving predictor values for condition-level forced-choice probabilities from an LLM such that, when fed into a common linking function (here with parameters $\alpha$ for optimization and $\epsilon$ for random error), the human choice probability can be reconstructed faithfully in its entirety.
On the other hand, there is a stark contrast between the results of item-level and condition-level data.
The model that was able to properly fit the condition-level data builds on predictions for item-level choice that are clearly incompatible with the human item-level data.
The model seems to be ``right for the wrong reasons.''

@@

@@

>>>>

To address this question, we assessed item-level scores $\text{S}_{ki}$ from the text-davinci-003 instance GPT-3.5 August 2023 \citep{BrownMann2020:Language-Models} for a text-based version of the reference game from Section~\ref{experiment-reference-games}.
% \footnote{
  % Numerically, we worked with \emph{average} log-probabilities. As in our experiment, all choice options had the same number of words, this is largely irrelevant, except for the effect of the $\alpha$ parameter in the interpretation condition. Using length-corrected log-probabilities is the same as using raw log-probabilities with an $\alpha$ parameter divided by the number of words. In this way, we make the $\alpha$ parameter more meaningfully comparable across conditions.}
The task description $x_{k}$ for item $I_{k}$ is a text supplied as prompt (see Appendix~\ref{sec:examples-items-llm} for an example).
The choice options are categorized as in the human experiment.\footnote{
  In the current set-up the response type ``distractor'' has two instantiations in the production condition. Since choice options are a single word in the production condition, for simplicity, we lump both of the distractor words together and treat them as a single option.}
Similar to the RSA model fit, we allow for random errors with probability $\epsilon_{C}$:
%
\begin{align*}
  P_{\text{item}}^{\text{LLM}}\left ( y_{ki} \mid C; \alpha_{c}, \epsilon_{c} \right )
  = \epsilon_{c} \  P_{\text{item}}^{\text{LLM}}\left ( y_{ki} \mid C; \alpha_{c}, \epsilon_{c} \right ) \ \ + \ \ \nicefrac{\epsilon_{c}}{3}   \,.
\end{align*}
%
The resulting likelihood function for item-level data with item-level predictions is, where priors on parameters are as for the RSA model described in Section~\ref{sec:model-pred-from}.:
%
\begin{align*}
 P^{\text{LLM}}_{\text{item}}(D_{c} \mid C; \alpha_{c}, \epsilon_{c}) = \prod_{k = 1}^{m} \text{Multinomial} \left ( D_{c}^{k}, \tuple{P^{\text{LLM}}_{\text{item}}(y_{ki} \mid C;  \alpha_{c}, \epsilon_{c})}_{1 \le l \le 3} \right )\,.
\end{align*}
%

@@

While LLMs do not make conditional-level predictions as such, they can be derived from item-level scores $\text{S}_{ki}$ by averaging over all items belonging to the relevant condition.
There are many ways of averaging item-level information.
Figure~\ref{fig:measures-overview} shows three salient approaches, which differ in what the underlying item-level measure is: the raw scores $\text{S}_{ki}$, the item-level probability derived from it (as used in Section~\ref{sec:item-level-pred}), or the predictions from the winner-takes-all (WTA) strategy commonly used in benchmark testing.

@@

The \emph{average-scores predictor} first aggregates the item-level scores, and then transposes the averages into (scaled) probabilities using the usual parameterized softmax function:\footnote{
  The reported results average over the multi-set \(\set{I_1, \dots I_m}\) of items that occurred in the human experiment for condition $C$.
  By using a multi-set, which may contain a single item multiple times, we produce aggregate predictions for exactly the set of items that the participant group saw, which provides the most fitting counterpart to the human data.}
%
\begin{align*}
  & P_{\text{cond}}^{\text{SCR}}(r_{i} \mid C, \alpha_{c})
    \propto \expo \left [  \alpha_{c} \ \frac{1}{m} \ \sum_{k=1}^{m} \text{S}_{ki}  \right ] \,.
    \tag*{\textcolor{gray}{[average scores (narrow-scope aggregation)]}}
\end{align*}
%
The \emph{average-probabilities predictor} first transposes scores into probabilities with a parameterized softmax function, and only aggregates over items last:
\begin{align*}
  % & \textbf{wide-scope aggregation} \\
  & P_{\text{cond}}^{\text{PRB}}(r_{i} \mid C, \alpha_{c})
    = \frac{1}{m} \ \sum_{k=1}^{m} P_{\text{item}}^{\text{LLM}}(y_{ki} \mid C, \alpha_{c}) \,.
    \tag*{\textcolor{gray}{[average probabilities (wide-scope aggregation)]}}
\end{align*}
%
Finally, the \emph{average-WTA predictor} considers the prediction of the WTA strategy (a softmax with $\alpha \rightarrow \infty$) as the basic item-level unit to aggregate over.
To add parameterized scaling to this method, a power-law transformation is a natural choice:
\begin{align*}
  % & \textbf{intermediate-scope aggregation} \\
  & P_{\text{cond}}^{\text{WTA}}(r_{i} \mid C, \alpha_{c})
    \propto  \left [ \frac{1}{m} \ \sum_{k=1}^{m} P_{\text{item}}^{\text{WTA}}(y_{ki} \mid C) \right ]^{\alpha_{c}}\,,
    \tag*{\textcolor{gray}{[average WTA (intermediate-scope aggregation)]}}
\end{align*}
where $P_{\text{item}}^{\text{WTA}}(y_{ki} \mid C) = \lim_{\alpha \rightarrow \infty} P_{\text{item}}^{\text{LLM}}(y_{ki} \mid C, \alpha)$.
\bigskip

The \emph{average-scores} and \emph{average-probabilities} predictors are equivalent if there is only one item, in which case the prediction of the \emph{average-WTA} method is the special case of $\alpha \rightarrow \infty$.
For cases with more than one item, the predictions of the three predictors are not guaranteed to be the same.
Conceptually, the \emph{average-probabilities} and the \emph{average-WTA} predictors, but not the \emph{average-scores} predictor, are compatible with a picture in which condition-level predictions result from the actual predictions at the item level.
However, based on the results from the previous section, the item-level predictions for the WTA strategy are demonstrably incongruent with the human data.
In general, the average-probability and average-WTA predictors can lead to qualitatively different results about task-level accuracy (see Appendix~\ref{sec:aver-prob-vs}).

\mf{UP TO HERE}

Using the same approach as for the RSA model in Section~\ref{sec:model-pred-from}, we can build likelihood functions for condition-level data around the three predictors introduced above.
With the same priors and methods used before, we obtain samples from the posterior over parameters and samples from the posterior predictive distributions.
Summary statistics for posteriors over model parameters are shown in Table~\ref{tab:sumStats}.
What is noteworthy is that for the \emph{average-WTA} model, the estimates of $\alpha_{c}$ in the production condition do not rule out, in fact lie close to, the special value $\alpha_{c}=1$, for which the power-law transformation is the identity function.
This means that just averaging WTA-responses at the item-level yields a reasonable predictor for the production data at the condition-level.
However, for the interpretation condition, the value $\alpha_{c}=1$ is clearly outside the range of credible parameter values, so that a simple recipe like ``always average WTA-responses without transformation'' is not a viable strategy for good condition-level predictions in general.
It is also worth noting that for the \emph{average-probability} model values of $\alpha_{c}$ of around five and above are virtually indistinguishable from the item-level predictions of the WTA strategy (see Figure~\ref{fig:closeness-target-item-level}).
This suggests that, for the production data, aggregation of item-level predictions from the WTA-strategy gives good condition-level predictions.

Figure~\ref{fig:refgame-counts} shows the summary statistics (means and 95\% credible intervals) for each model's posterior predictive distribution.
We find that only the theoretical model (RSA) and the \emph{average-WTA} model pass this ``visual posterior predictive check'' for both conditions; the other two models both overpredict the target choice rate and underpredict the competitor choice rate in the interpretation condition.
To corroborate the visual impression, Table~\ref{tab:sumStats} shows sample-based estimates of Bayesian posterior predictive $p$-values, using likelihood of the observed data as a test statistics.
Consequently, the results from Table~\ref{tab:sumStats} suggest that the \emph{average-scores} and the \emph{average-probabilities} model are able to reproduce the production data, but fail on the interpretation data; and that only the \emph{average-WTA} model does not fail to capture the data from both conditions.

These results tell us that not all ways of deriving condition-level predictions by averaging over item-level variation are equally good.
Some approaches clearly fail basic checks for statistical goodness-of-fit.
On the positive side, we also find that there is at least one model with predictors based on LLM-measures, namely the \emph{average-WTA} model, which is able to recover the patterns in the data.
In other words, there is a way of deriving predictor values for condition-level forced-choice probabilities from an LLM such that, when fed into a common linking function (here with parameters $\alpha$ for optimization and $\epsilon$ for random error), the human choice probability can be reconstructed faithfully in its entirety.
On the other hand, there is a stark contrast between the results of item-level and condition-level data.
The model that was able to properly fit the condition-level data builds on predictions for item-level choices that are clearly incompatible with the human item-level data.
The model seems to be ``right for the wrong reasons.''

@@

All previous results where obtained for LMM predictions derived from GPT-3.5 turbo.
To investigate whether key results replicate for other LLMs and scales, we ran the exact same analyses for item- and condition-level data also for scores derived from variants of LLaMA2 \citep{TouvronLavril2023:LLaMA:-Open-and}, in particular the 7, 13 and 70 billion parameter versions of the versions fine-tuned with human feedback only (marked as `hf' in the following) and based on human feedback and chat data (marked as `chat').
Table~\ref{tab:sumStats-LLaMA2} shows the relevant summary statistics and Bayesian posterior predictive $p$-values.
Figure~\ref{fig:LLaMA-results-vPPC} shows visual posterior predictive checks for the condition-level data.
The plots show the differences between human counts for each condition and option and the models' posterior predictions.
We find that the average-WTA predictor is consistently able to capture the data from the production condition except for one of the six models (LLaMA2-chat 70B).
However, the average-WTA predictor only recovers the interpretation data for only one of the six models (LLaMA2-chat 13B).
Interestingly, for this LLM as backend for scoring, also the average-probability predictor is able to recover the interpretation data, unlike for the GPT-3.5 backend.

\mf{TODO: item-level; check estimated alphas for condition-level!?}

In sum, we find general support for the previous conclusion that the average-WTA predictor is most successful in capturing the condition-level data, but we also find variability in which method of aggregation works well with which LLM as scoring model.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Average-probability vs.~average-WTA for accuracy scores}
\label{sec:aver-prob-vs}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Standard benchmark testing looks at \emph{task accuracy}, defined as the probability of selecting the ``gold standard'' response, usually based on a WTA selection of the highest scoring option.
We can generalize this and define the \emph{softmax-based accuracy}, $P_{\text{cond}}^{\text{SM}}\left(r_{1} \right)$, as the average probability of choosing the designated target response option $r_{1}$ for the \emph{condition-level soft-max prediction}:
\begin{align*}
  P_{\text{cond}}^{\text{SM}}\left(r_{i} \right) = \frac{1}{m} \sum_{k = 1}^{m} P_{\text{item}}^{\text{SM}} \left(y_{ki} \right)\,.
\end{align*}
As before, the average-WTA task accuracy is the special case of $\alpha \rightarrow\infty$.

Indeed, WTA-based accuracy can differ qualitatively from softmax-based accuracy, as shown by the following example.
The example shows that differences between accuracy measures depend on the variation in item-level scores, in particular on the relation between score-ordering and score-differences.
Note that the example holds equally if numbers for the two options are reversed, so that there is no way of saying which of the two measures of accuracy would generally be more favorable for selecting the target option.
%
\begin{quote}
  % \ex \label{bsp:exmpl-diff-WTA-SM}
  \textbf{Example:}
  Imagine that there are two options, and that the target option's score is a small $\epsilon$ higher in 80\% of the task's items, and otherwise lower.
  The WTA-based accuracy is 0.8.
  This number is useful as a performance measure for applications in which the LLM is used in exactly the way the WTA strategy describes, e.g., any implementation which is outcome equivalent to greedy decoding with rejection sampling on a domain that contains only the available options.
  For such a case, it never matters how much worse the goal answer is scored in the 20\% of the cases where it is not the maximum.
  As only the best option will be chosen, that information is irrelevant.
  But if an application uses anything other than greedy-like responses, the accuracy score of 0.8 may be misleading.
  If the remaining 20\% of the items are such that the non-goal option is almost infinitely better, it would be chosen under a pure sampling strategy, where $\alpha = 1$, with virtual certainty, so the softmax-based accuracy would around 0.4.\footnote{The probability of the target option in the 80\% of items where the goal answer is slightly better is 0.5 in the limit of $\epsilon \rightarrow 0$, and it is virtually 0 in the remaining 20\% of the cases. This gives an expected rate of: $\nicefrac{4}{5} \ \nicefrac{1}{2} + \nicefrac{1}{5} \ 0 = \nicefrac{2}{5}$.}
\end{quote}
%
The upshot of these considerations is that the standard practice of WTA-based performance assessment for LLMs gives false, or at least misleading or inaccurate results, whenever not all downstream applications use a greedy-like sampling strategy (which is almost certainly the case), and there is variability in item-level predictions.
